# Introduction {#intro}

```{r echo = FALSE, warning = FALSE, error = FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
```


*Streptococcus pneumoniae* is a commensal bacterium found in the nasopharynx of humans where it plays an integral role in the normal upper respiratory flora.
It is also a common pathogen and is one of the most common bacterial causes of disease in humans.
In classical medical texts pneumococcus is described as a Gram-positive lancet-shaped coccus that is usually found in pairs.
In fact, pneumococcus is *the* Gram-positive coccus, being first bacteria that Christian Gram noted which retained the dark aniline-gentian violet stain that now bears his name [@Gram1884].
Pneumococcus was first isolated in 1881 by two microbiologist, George M. Sternberg in the United States and Louis Pasteur in France [@Pasteur1881; @Sternberg1881; @Watson1993].
In both cases the sample from which the bacterium was isolated originated from healthy carriers, rather than ill patients.
The causal association between this newly discovered bacterium and pneumonia was firmly established only five years later [@Weichselbaum1886], and in the following decade, all clinical presentations of pneumococcal infection had been described [@Austrian1981].
Pneumococcus has gone by many names since its first isolation in 1881.
Originally it was named *Micrococcus pasteuri* by Sternberg [@Sternberg1881] but, by 1920 a scientific consensus was reached in which it was agreed that the official name should be *Diplococcus pneumoniae* [@Winslow1920].
It was not until 1974 that pneumococcus received its current name, *Streptococcus pneumoniae* [@Deibel1974].

Pneumococcus is encapsulated by a polysaccharide coating that protects it from environmental factors.
The polysaccharide capsule acts as an "invisibility cloak" to the human immune system, which is rendered unable to detect pneumococcus except through certain patterns in the oligosaccharides contained within the capsule [@Epstein1995].
Based on these patterns, pneumococcus has been classified into over 97 different serotypes to date.
Additionally, as the capsule contains only polysaccharides and not proteins, the immune response is T-cell independent and therefore poorly immunogenic, even after identification by the immune system [@Geno2015b].
These characteristics are the fundamental challenges faced by scientists engineering new pneumococcal vaccines.
The significance of serotypes on the development of vaccines should not be understated.
The failure of the original attempt of Wright and colleagues in 1911 to prevent pneumococcal pneumonia with vaccination was indeed due to lack of knowledge about serotype-specific immunogenicity [@Wright1914].

All serotypes of pneumococcus have the potential to cause disease in humans.
However, some are more virulent than others.
The prevalence of asymptomatic carriage in the nasopharynx varies greatly by serotype as does the propensity of serotypes to cause clinical infections.
Quantifying the pathogenic potential of serotypes is difficult as both their prevalence and propensity to cause disease need to be considered. With few exceptions, acquisition of a new serotype into the nasopharyngeal flora proceeds the onset of clinical disease caused by that serotype.
Pneumococcal epidemiology is dominated by this effect - children act as a reservoir of asymptomatic pneumococcal carriage from which other children and adults acquire serotypes that may lead to symptomatic disease. 
Because of this, vaccinations that decrease the pneumococcal carriage in children have the potential of reducing the incidence of disease both in other unvaccinated children and in adults. 
This phenomenon is called a herd-effect, and is integral to the development of vaccination strategies to combat pneumococcal disease.

The infectious manifestations of pneumococcal disease are, broadly speaking, local infections of the respiratory tract and infections of previously sterile tissue.
They range from common to uncommon, and from benign to serious.
The most common infectious manifestation of pneumococcus in acute otitis media (AOM) – an infection of the middle ear.
AOM is the most common reason for physician visit and for antimicrobial prescription in the paediatric population.
However, the disease course is benign and rarely results in permanent disability.
The pathogenesis of AOM is complex and can both be caused by viral and bacterial pathogens.
Bacterial AOM is most often caused by pneumococcus and *Haemophilus influenzae*.
Acute sinusitis is another common, though benign, manifestation of pneumococcal disease.
Pneumococcus is also a common cause of pneumonia – the disease from which it gets its name.
Pneumonia often requires hospitalization and intravenous antimicrobial treatment, and, though uncommonly, can lead to permanent disability and death.
Finally, if pneumococcus gains access to normally sterile tissue, it may cause invasive infections.
This includes bacteremia, an infection of the blood, and meningitis, an infection of the meninges.
These infectious manifestations are grouped together as invasive pneumococcal disease (IPD).
Whilst IPD is extremely uncommon, the consequences can be disastrous.
The case fatality ratio from pneumococcal meningitis in Iceland is estimated at 15.3%. 

Pneumococcus became an early target for vaccine development because of the broad range of disease caused by pneumococcus.
The vaccine benefit can be quantified in two different ways.
On one hand it can prevent uncommon but serious manifestations of disease and has the potential to prevent death and serious disability.
On the other hand, it also prevents common infections that cumulatively present a large healthcare burden due to frequent physician visits, antimicrobial prescriptions as well as work hours lost by parents and caregivers.
The earliest attempts to use vaccination to lessen the morbidity and mortality associated with pneumococcus date back to 1911 when Wright used whole cell innoculi to vaccinated miners in South Africa [@Wright1914].
In the following decades, multiple animal studies showed that injection of the pneumococcal polysaccharide coating in animals protected against subsequent pneumococcal infections.
On the basis of these findings, the first polysaccharide vaccine was shown to be effective [@Macleod1945].
This lay the foundation for modern polysaccharide pneumococcal vaccines. 
However, it soon became apparent that the polysaccharide vaccines were not adequately immunogenic in young children, the ill or the elderly.
In response, pneumococcal conjugate vaccines were developed.
The first such vaccine to be mass produced was the heptavalent Prevenar developed by Pfizer Pharmaceuticals.

Iceland is an independent island nation, isolated in the mid-Atlantic, with a homogenous populuation of roughly 330,000 individuals.
The first systematic program of vaccination against pneumococcus in Iceland began in April 2011, when the 10-valent pneumococcal *Haemophilus influnzae* protein-D conjugate vaccine (Synflorix, PHiD-CV10) was introduced into the national paediatric vaccination program.
The vaccine program entailed two primary doses given at three and five months of age, and a booster dose at twelve months.
No catch-up program was undertaken.
Prior to the introduction, no systematic vaccination effort had been undertaken in Iceland.
Iceland, as other Nordic countries, has a rich legacy of national health-related registers.
With a wealth of medical documentation, a unique whole-population ecological study examining the impact of systematic pneumococcal vaccination was enabled.


## Clinical manifestations of *Streptococcus pneumoniae*

The relationship between pneumococcus and humans is complex. 
Most children are colonized by pneumococcus within the first months of life. 
The serotype distribution of the initial colonization is influenced by the distribution of serotypes within the family. 
Over the course of the child's lifetime, they will be colonized by many different serotypes.
The child's immune system will learn to recognize the currently colonizing serotypes and will either clear the colonization or maintain a equilibrium in which the serotype is kept within a certain limit of reproduction.
Consequently, the contribution of pneumococcus to the human upper respiratory flora is in a state of constant flux.
New serotypes enter while old exit, and the relative density of serotypes changes.
In some cases the equilibrium between pneumococcus and the host is destabilized resulting in rapid growth of pneumococcus which results in clinical manifestations.
It is thought that this is most likely to occur immidietly following the acquisition of new serotype into the nasopharyngeal flora. 
This most commonly occurs in the upper respiratory tract where pneumococcus is generally located and results in the common clinical manifestations of pneumococcal infections, i.e. AOM, acute sinusitis and conjunctivitis.
The pathogenesis of pneumococcal pneumonia is thought to occur through micro-aspiration of upper respiratory secretions with subsequent rapid proliferation of pneumococcus in the lower respiratory tract.
Invasive disease occurs when pneumococcus penetrates the host immunological defenses and proliferates in normally sterile tissue.
This can be secondary to infections of the upper or lower respiratory tract, or can occur as a primary event.
Generally, IPD is considered to encompass meningitis, bacteraemia and septic arthritis.
While some may argue that the middle ear is normally sterile, AOM is not considered invasive disease.

It has been known from the first pneumococcal vaccine trials that vaccination has different efficacy against the different manifestations of pneumococcal disease.
The largest effects are consistently seen in the prevention of IPD and pneumococcal lobar pneumonia.
The effects on carriage and AOM are often lesser in magnitude. 
This may be a true biological gradient or a consequence of the accuracy with which disease is measured. 
The more serious the illness is the more testing is performed resulting in a more accurate diagnosis.
Much of the AOM attributed to pneumococcus may be caused by other pathogens, while by definition IPD is always caused by pneumococcus.
Furthermore, it wasn't until the advent of pneumococcal conjugate vaccines that vaccines started to become efficacious in children and it is precisely in children that AOM occurs. 
The largest trials of modern pneumococcal vaccines have fit the above narrative.
The 23-valent pneumococcal polysaccharide vaccine was trialed in 12,000 adults and showed an efficacy of 75% in preventing IPD and a 50% efficacy against radiologically confirmed pneumonia.
The heptavalent pneumococcal conjugate vaccine was trialed in children and produced a 97% efficacy in preventing IPD. 



### Carriage

~ 3-4 pages
- Define carriage; age-specific prevalence, serotype distribution
	- Explain that most are born carriage free
	- Evidence for co-carriage of different serotypes
	- Age at which most children acquire carriage
		- Risk factor: daycare, siblings, smoking etc.
	- Children are the main vectors of pneumococcus
	- Rate of clearance dependent on age
	- With increasing age -> increasing immunity, decreasing prevalence
	- Senescence and carriage in the elderly
- Evidence for carriage being the predecessor infections
	- Evidence of asymptomatic carriage -> main spread of disease
	- Variable propensity of serotypes to cause disease, attack-rates
- Review the Icelandic literature and changing epidemiology
	- Carriage prevalence
	- Serotype distribution
	- Risk factors

### Acute otitis media

~ 3 - 4 pages
- Define different types of otitis media; acute otitis media
	- Pathogens, estimated % caused by pneumococcus
	- Proposed mechanism by which carriage -> AOM
	- Epidemiology, both serotype and age
	- Risk factors
- Burden of disease caused by AOM; health care utilization, cost
	- Incidence and prevalence
	- GP visits, antibacterial consumption, hospitalization (?)
	- Days of work-lost by parents
	- Sequelae; multiple infections, effusion, tympanostomies 
	- Evidence of benefit of delaying 1st presentation
-  Review of Icelandic literature and changing epidemiology
	- AOM prevalence and serotype distribution
	- Risk factors
	- Associated healthcare consumption, cost
	- Rate of sequelae

### Pneumonia

~ 4 – 5 pages
- Define: CAP, nosocomial, PP, NBPP and IPP. 
	- Pathogens, estimated % caused by pneumococcus
	- Proposed mechanism by which carriage -> pneumonia
	- Epidemiology, both serotype and age
	- Risk factors
- Burden of disease caused by pneumonia, health care utilization
	- Ways of defining severity; CURB-65 etc. 
	- GP visits, antibacterial consumption, Hospitalization rates
	- Days of work lost	
	- Mortality, sequelae
-  Review of Icelandic literature and changing epidemiology
	- Pneumococcal pneumonia prevalence and serotype distribution
	- Rate of hospitalization, healthcare consumption
	- Rate of sequelae
	- Risk factors

### Invasive pneumococcal disease

~ 3 -5 pages
- Define different presentations of IPD: meningitis, bacteremia, etc.
	- Epidemiology, both serotype and age
	- Risk factors
	- Burden of disease, health care utilization
		- Severity
		- Hospitalization rates, ICU rates
		- Sequelae
-  Review of Icelandic literature and changing epidemiology
	- Meningitis, bacteremia, empyema, joint infection prevalence and 		   serotype distribution
	- Rate of sequelae


## Pneumococcal conjugate vaccines

The history of pneumococcal vaccination begins in 1911 when Wright and colleagues attempted to use whole killed bacteria to vaccinate South African miners against pneumococcal pneumonia [@Wright1914].
It should however be noted that in his original description of pneumococcus in 1881, George Sternberg observed that rabbits who were injected with his saliva mixed with alcohol and quinine did not ubiquitously die and were later resistant to re-injection with saliva [@Austrian1999a, @Sternberg1881].
Sternberg had inadvertently immunized the laboratory animals against subsequent infection by injecting killed pneumococci, proving the concept 30 years before it was first attempted.
The 1911 trial by Wright failed to show efficacy due to lack of knowledge of the significance of serotypes and serotype specific immunogenicity.
Several trials using whole killed bacteria were published in the following two decades [@Cecil1918; @Lister1916; @Lister1936; @Maynard1913]
The researchers benefited from the knowledge of serotypes which had been discovered in 1910 and used multivalent vaccines.
They were however victims of underdeveloped epidemiological methodology for vaccine field trials.
Due to inconsistencies in study design, the efficacy of whole bacteria pneumococcal vaccines remained fiercely debated though there seemed to be a suggestion of benefit [@Austrian1999a].
Vaccines based on whole killed bacteria were soon replaced with polysaccharide vaccines, following discoveries in the 1920s and 1930 of the immunogenicity of the polysaccharide capsule [@Dochez1917; @Finland1931; @Francis1930; @Heidelberger1923; @Schiemann1927].
The first such trial tested a bivalent polysaccharide vaccine on 29,000 adult males in the American Civilian Conservation Corps in iterations in the 1930s [@Ekwurzel1938].
It suffered from the same methodological problems as did the previous trials of the whole killed bacteria and its results were debated.
A second large trial was conducted in the late 1930s using a tetravalent vaccine [@Macleod1945].
This trial built upon the experience of the previous trials and was able to show convincing efficacy against pneumococcal pneumonia.
The results of this trial led to the licensure of two hexavalent polysaccharide pneumococcal vaccines in the 1940s.
One was formulated for adults and the other for children, each optimized to the serotype distribution within the respective age-group.
Alas, these early vaccines fell victim to unfortunate timing.
Because in 1944, Tillet and colleagues showed that bacteraemic pneumococcal pneumonia could be cured by parenteral administration of benzylpenicillin [@Tillett1943].
Following this discovery, the medical community was stricken with a kind of nonchalance.
The mortality rate of pneumococcal disease decreased sufficiently that there was no longer a perceived need for preventative vaccination.
The licenses for the polysaccharide vaccines were withdrawn by the manufacturer due to lack of their use [@Austrian1999a]
Interest in pneumococcal vaccination re-emerged in the 1950s when it was noted that the mortality benefit of penicillin was not ubiquitous.
The elderly and those who had underlying disease did not experience a decrease in their case fatality ratio [@Austrian1964].
This led to a redoubled effort to create a new polysaccharide vaccine.
Several large randomized controlled trials were conducted in South Africa in the 1970s [@Austrian1976, @Smit1977] and on the basis of these, a 14-valent pneumococcal vaccine was licensed in the United States in 1977.
Its valency was increased to 23 polysaccharides in 1983 [@Austrian1999a].
Early in the development of pneumococcal vaccinations there was an interested in vaccinating children.
Two trials were conducted in the early 1980s which attempted to use polysaccharide vaccines in young children. 
Neither showed benefit [@Makela1981; @Sloyer1981].
This is perhaps unsurprising in light of previous trials.
The first polysaccharide trial that was conducted in children in 1937 failed to detect any immunological response [@Davies1937].
Laboratory studies in the 1930s and 1940s revealed that the reason for this lack of efficacy was due to the thymus independent immunse response to purely sacharide antigens.
These same studies showed that this could be remedied by adding a protein adjuvant, thus inducing a T-cell response.
The strategy of protein conjugation saw its first success in the development of the *Haemophilus influenzae* type b vaccine.
Subsequently, several different pneumococcal conjugate vaccines entered phase II and phase III clinical trials in the late 1990s [@Austrian1999a].
The first of which to recieve licensure was the 7-valent pneumococcal conjugate vaccine (PCV7) which was licensed in 2000 in the United States.
It included purified polysaccharides for seven serotypes of pneumococcus (4, 9V, 14, 19F, 23F, 18C and 6B) conjugated to CRM197, a nontoxic variant of the diptheria toxin.
It was shown to be efficacious for IPD, pneumococcal pneumonia and AOM in several randomized trials [@Black2000].
Higher valancy conjugated vaccined were developed and recived licensure in the new millenium based on the randomized trials conducted for the heptavalent conjugated vaccine.
They have however been shown to be effective in several cluster randomized trials and observational studies. 


## Impact of pneumococcal conjugate vaccines

~ 3 pages
- Present evidence of magnitude of effect on VT carriage
	- Serotype distribution vs. carriage prevalence
		- Serotype replacement 
	- Herd-effect, i.e. effect on carriage of adults and non-vaccinated

From their inception, dozens of randomised controlled trials evaluating the efficacy of differenct pneumococcal conjugate vaccines have been performed.



### Acute otitis media

~ 1-2 pages <- much fewer studies
- Present evidence of magnitude of effect on all-cause AOM
	- VT vs. NVT serotypes
		- Serotype replacement (?)	
	- Herd-effect in non-vaccinated
	
Acute otitis media is still most often caused by *Streptococcus pneumoniae* and *Haemophilus influenzae* despite changes in otopathogens.
Prevention of IPD in children and the associated morbidity and mortality was the driving force in the development of pneumococcal conjugate vaccines.
However, the public most often associates them with AOM.
Most children experience AOM and the dramatic decrease in incidence following pneumococcal conjugate vaccination is what families noticed.
Despite this, AOM is a difficult outcome for trialist.
AOM exists on a continuum.
It does not have universally adhered to diagnostic criteria and its signs and symptoms greatly overlap with those of other common diseases. 
Because AOM is benign and most often self-limited, the probability that a child with AOM is even seen by a physician varies greatly with parental health seeking behavior.
Even when AOM is accurately diagnosed it is not possible to ascertain the causative pathogen without invasive sampling, which is not warranted given the benign nature of the disease.
This percludes measuring the serotype specific effect of vaccination for most studies - and more importantly, it percludes measuring the effect on pneumococcal AOM. 
Thus any estimation of an effect of pneumococcal vaccination will necessarily by diluted by the subjectiveness of AOM diagnosis and the continued lack of protection against other otopathogens. 

Despite these difficulties, AOM has been associated with pneumococcal vaccination in children from the beginning.
It was used as an outcome measure in the earliest trials of the pneumococcal polysaccharide vaccines [@Makela1981; @Sloyer1981].
The first published randomized controlled trial of a pneumococcal conjugate vaccine reported, among other outcomes, the efficacy against AOM [@Black2000].
The study obtained data on visits for AOM from routine electronic health records.
The assessors were not specifically trained as these were simply routine visits. 
The outcome measure AOM was defined in at least eight different ways to account for the difficulties in measurement.
Only per protocol effects will be presented here though none of the intention to treat results were divergent from them.
The vaccine efficacy against otitis media 

```{r Table 1, echo = F}
kable(
    data.frame(
        study = c("Black, 2000", "Eskola, 2001"),
        vaccine = c("PCV7 (CRM197)", "PCV7 (CRM197)"),
        no_child = c(37868, 1662),
        efficacy = c("8.9 (5.8-11.8", "6 (-4-16)")
    ),
    col.names = c("Study", "Vaccine", "No. of children", "Efficacy against Otitis media episodes"),
    booktab = T
)
```




### Pneumonia

~ 2-3 pages
- Present evidence of effect on all-cause pneumonia
	- VT vs. NVT serotypes
		- Serotype replacement (?)
	 - Herd-effect in adults and non-vaccinated


### Invasive pneumococcal disease

~ 4-6 pages <- largest amount of studies
- Present evidence of effect on IPD and subgroups; meningitis, bacteremia etc. 
	- VT vs. NVT
	- Serotype replacement
	- Herd-effect

## Cost-effectiveness of pneumococcal conjugate vaccination

~ 3-4 pages
- Present overview of literature review and critical analysis. 
- Recommendations of ISPOR and WHO presented, discuss importance of assumptions and methodology
- Introduction to sub-chapters of lit. rev. 
	- Explain how they will be tied in to ISPOR/WHO recommendations


### Measurement of effectiveness and choice of health outcomes

~ 1 page
- Shortly explain what is meant by effectiveness and health outcomes
	- Tie in to ISPOR/WHO

#### Health outcomes considered

~ 2-3 pages
- Describe what health outcomes were considered
	- Tie into actual disease burden known to be caused by pneumococcus
	
#### Effectiveness of PCV7

~ 3-4 pages
- What effectiveness rationale is used, methods and rationale: critique
		- Carriage
		- AOM
		- Pneumonia
		- IPD


#### Effectiveness of PCV10

~2-3 pages
- What effectiveness rationale is used, methods and rationale: critique
		- Carriage
		- AOM
		- Pneumonia
		- IPD

#### Effectiveness of PCV13

~ 2- 3 pages
- What effectiveness rationale is used, methods and rationale: critique
		- Carriage
		- AOM
		- Pneumonia
		- IPD


### Estimating resources and cost


~1 page
- Shortly explain what resources and costs mean
	- Direct vs. indirect
	- Tie in to ISPOR/WHO
