# Materials and methods {#methods}

## Data collection and sources {#datasources}

During the study period from 1 January 2005 to 31 December 2017, data were collected from multiple whole population registries, and from the patient registry of Landspitali University Hospital.
Data were collected based on government issued national identification numbers.
Each individual receives only one number over the course of their lifetime, and the identification number is permanently retired at the time of death.
The Icelandic Directorate of Health processed and anonymized all data from the various registries before releasing it to the study group. 
A study identifier was created based on the national identification number, which was then removed from the data as part of the anonymization process. 
The mapping key was kept by the Directorate of Health, and was not accessible to the study group.
The study group linked data from the various registries using both the study identifier and dates of events. 

In the following sub-chapters, each registry providing study data is reviewed.
Statistics Iceland provided data on immigration and emigration, demographic indices and salaries.
Diagnostic data were obtained from Landspitali University Hospital's patient registry and the Primary Care Registry of the Directorate of Health. 
Pneumococcal vaccination status was collected from the National Vaccine Registry (NVR) and augmented with information on privately purchased vaccine doses obtained from the National Drug Prescription Registry (NDPR).
Data regarding antimicrobial prescriptions were also extracted from the NDPR.
Finally, reimbursement data for outpatient otolaryngological procedures were obtained from Icelandic Health Insurance.

### Statistics Iceland {#statisticsiceland}

Statistics Iceland collects and maintains a large array of economic, social and demographic indices, and provides aggregate data at www.statice.is.
For each calendar-year 2005-2017, the number of individuals living in Iceland was collected from Statistics Iceland, stratified by postal-code, gender and age in years.
These data were used for the denominator in incidence calculations in all papers.
The deciles of salary from 2005-2017 were obtained from Statistics Iceland and used to inform a sensitivity analysis on the cost-effectiveness of PHiD-CV10 ( [Paper VI](#paper6)).
Costs were adjusted for inflation using the Medical Care Consumer Price Index of Statistics Iceland, and wages adjusted using the National Wage Index. 
In addition to the aggregate data presented above, individual-level information on the immigration and emigration of children zero to four years of age was obtained, anonymized and linked to the other study data.  

### Landspitali University Hospital patient registry {#patientregistry}

Landspitali University Hospital is the sole tertiary hospital in Iceland, and includes Children's Hospital Iceland -- Iceland's only pediatric hospital. 
It provides primary and secondary care for the capital area, approximately 65% of the Icelandic population, and tertiary care for the whole population.
In 2017, the total number of non-psychiatric curative care hospital beds in Iceland was 732 (www.statice.is).
Of those, 669 (91%) were at Landspitali University Hospital.
Landspitali's patient registry records information on all emergency department and outpatient visits, and all hospital admissions to Landspitali University Hospital. 
For the period from 1 January 2005 to 31 December 2017, data were extracted on all unplanned acute-care visits and hospital admissions with International Classification of Diseases, 10th revision (ICD-10) discharge diagnoses compatible with respiratory infections (Table \@ref(tab:table-icd)).

```{r table-icd}
kable(
    x = data.frame(
    stringsAsFactors = FALSE,
    icd = c(
    "A40",
    "A41",
    "A48",
    "A49",
    "B00",
    "B08",
    "B33",
    "B34",
    "B95",
    "B96",
    "G00",
    "H65",
    "H66",
    "H70",
    "H72",
    "H73",
    "J00",
    "J01",
    "J02",
    "J03",
    "J04",
    "J05",
    "J06",
    "J09",
    "J10",
    "J11",
    "J12",
    "J13",
    "J14",
    "J15",
    "J16",
    "J17",
    "J18",
    "J20",
    "J21",
    "J22",
    "J32",
    "J36",
    "J40",
    "J85",
    "J86",
    "J90",
    "N30",
    "N39",
    "R05",
    "R50"
    ),
    Disease = c(
    "Streptococcal sepsis",
    "Other sepsis",
    "Other bacterial diseases, not elsewhere classified",
    "Bacterial infection of unspecified site",
    "Herpesviral [herpes simplex] infections",
    "Other viral infections characterized by skin and mucous membrane lesions, not elsewhere classified",
    "Other viral diseases, not elsewhere classified",
    "Viral infection of unspecified site",
    "Streptococcus, Staphylococcus, and Enterococcus as the cause of diseases classified elsewhere",
    "Other bacterial agents as the cause of diseases classified elsewhere",
    "Bacterial meningitis,not elsewhere classified",
    "Nonsuppurative otitis media",
    "Suppurative and unspecified otitis media",
    "Mastoiditis and related conditions",
    "Perforation of tympanic membrane",
    "Other disorders of tympanic membrane",
    "Acute nasopharyngitis [common cold]",
    "Acute sinusitis",
    "Acute pharyngitis",
    "Acute tonsillitis",
    "Acute laryngitis and tracheitis",
    "Acute obstructive laryngitis [croup] and epiglottitis",
    "Acute upper respiratory infections of multiple and unspecified sites",
    "Influenza due to certain identified influenza viruses",
    "Influenza due to other identified influenza virus",
    "Influenza due to unidentified influenza virus",
    "Viral pneumonia, not elsewhere classified",
    "Pneumonia due to Streptococcus pneumoniae",
    "Pneumonia due to Hemophilus influenzae",
    "Bacterial pneumonia, not elsewhere classified",
    "Pneumonia due to other infectious organisms, not elsewhere classified",
    "Pneumonia in diseases classified elsewhere",
    "Pneumonia, unspecified organism",
    "Acute bronchitis",
    "Acute bronchiolitis",
    "Unspecified acute lower respiratory infection",
    "Chronic sinusitis",
    "Peritonsillar abscess",
    "Bronchitis, not specified as acute or chronic",
    "Abscess of lung and mediastinum",
    "Pyothorax",
    "Pleural effusion, not elsewhere classified",
    "Cystitis",
    "Other disorders of urinary system",
    "Cough",
    "Fever of other and unknown origin"
    )
    ),
    col.names = c("ICD-10 code", "Disease"),
    caption = "The International Classification of Diseases, 10th revision (ICD-10) codes used for individual-level data collection from the Primary Care Registry and Landspitali University Hospital's patient registry. All subgroups of the listed ICD-10 codes were also included."
    )
```

Additionally, any visit or hospital admission associated with NOMESCO Classification of Surgical Procedures (NCSP) procedural codes in Table \@ref(tab:table-ncsp) were extracted the patient registry. 

```{r table-ncsp}
kable(
    x = data.frame(
    stringsAsFactors = FALSE,
    NCSP.code = c(
    "EMSB00",
    "EMSB10",
    "EMSB15",
    "EMSB20",
    "EMSB30",
    "EMSB99",
    "EMSW99",
    "DCSA10",
    "DCSA20",
    "DCSW00"
    ),
    Description = c(
    "Excision of lesion of tonsil or adenoid",
    "Tonsillectomy",
    "Intracapsular destruction of tonsils",
    "Adenotonsillectomy",
    "Adenotomy",
    "Other excision on tonsils and adenoids",
    "Other operation on tonsil or adenoids",
    "Paracentesis of tympanic membrane",
    "Insertion of ventilating tube through tympanic membrane",
    "Removal of ventilating tube from tympanic membrane"
    )
    ),
    col.names = c("NCSP code", "Description"),
    caption = "NOMESCO Classification of Surgical Procedures (NCSP) codes used for individual-level data collection from Landspitali University Hospital's patient registry."
    )
```


The data included the date of visit or hospital admission, date of hospital discharge, hospital length of stay, the departments involved (including the intensive care unit), and a detailed breakdown of costs associated with each contact.
A separate and unique identification number was provided for each individual visit or hospital admission.
All costs were recorded in Icelandic kronas (ISK) and were broken down into specific subsets.
Costs associated with diagnostic testing were provided and categorized as costs associated with chemical blood testing; diagnostic radiological testing; anatomical pathology; virological testing; bacteriological cultures; antibody and other immunological testing; and specific tests performed by the blood bank in preparation for the administration of blood products.
Costs associated with departmental upkeep, such as heat, electricity, and wages were divided between patients based on hospital length of stay.
The costs associated with treatment were divided into the cost of drugs, surgery and procedures and intensive care unit treatment. 

Several smaller independent data-sets pertaining to specific papers were extracted from the patient registry.
These data-sets were not linked to the main study data. 

In [paper I](#paper1), describing the impact of PHiD-CV10 on acute otitis media with treatment failure, information on all doses of ceftriaxone administered at the Children's Hospital Iceland between January 2009 and December 2015 was extracted from the hospital's medication administration system using the ATC code J01DD0.
Any ICD-10 diagnostic code associated with a visit or hospital admission in which ceftriaxone was administered, was extracted from the patient registry.
Importantly, this included all ICD-10 codes, not only those in Table \@ref(tab:table-icd).
Also obtained for [paper I](#paper1) was the aggregate number of yearly visits to the pediatric emergency department of Children's Hospital Iceland 2008-2015.

[Paper VI](#paper6) -- a cost-effectiveness analysis of PHiD-CV10 introduction into the pediatric vaccination program, required control diseases used within a time-series analysis framework. 
The aggregate monthly number of acute-care visits and hospital admissions for several sub-chapters of the ICD-10 diagnostic coding system (Table \@ref(tab:table-icd-synthetic)) was obtained for 22 different age-groups. 

```{r table-icd-synthetic}
kable(
    x = data.frame(
    stringsAsFactors = FALSE,
    icd.code = c(
    "A10-B99",
    "C00-D48",
    "D50-89",
    "E00-99",
    "G00-G99",
    "H00-99",
    "I00-99",
    "K00-99",
    "L00-99",
    "M00-99",
    "N00-99",
    "P00-99",
    "Q00-99",
    "R00-99",
    "S00-T99",
    "U00-99",
    "V00-Y99",
    "Z00-99"
    ),
    Description = c(
    "Certain infectious and parasitic diseases",
    "Neoplasms",
    "Diseases of the blood and blood-forming organs and certain disorders involving the immune mechanism",
    "Endocrine, nutritional and metabolic diseases",
    "Diseases of the nervous system",
    "Diseases of the eye and adnexa, Diseases of the ear and mastoid process",
    "Diseases of the circulatory system",
    "Diseases of the digestive system",
    "Diseases of the skin and subcutaneous tissue",
    "Diseases of the musculoskeletal system and connective tissue",
    "Diseases of the genitourinary system",
    "Certain conditions originating in the perinatal period",
    "Congenital malformations, deformations and chromosomal abnormalities",
    "Symptoms, signs and abnormal clinical and laboratory findings, not elsewhere classified",
    "Provisional assignment of new diseases of uncertain etiology",
    "Injury, poisoning and certain other consequences of external causes",
    "External causes of morbidity",
    "Factors influencing health status and contact with health services"
    )
    ),
    col.names = c("ICD-10 code", "Description"),
    caption = "The International Classification of Diseases, 10th revision subchapters used to define the synthetic controls used in time-series analyses."
    )
    
```


Data from the patient registry were used in [paper I](#paper1), [paper IV](#paper4), [paper V](#paper5) and [paper VI](#paper6).

### The Primary Care Registry {#primarycareregistry}

In the Icelandic healthcare system, primary care is provided by family medicine physicians at 69 neighborhood based centers (*Heilsugæsla*).
All primary care centers use the same electronic medical record system, and the same diagnostic coding systems (NCSP, ICD-10) as Landspitali University Hospital and Children's Hospital Iceland.
The Directorate of Health maintains a registry on all primary care visits within the Icelandic healthcare system.
From this registry, all physician visits with ICD-10 diagnostic codes compatible with respiratory tract infections were extracted for the period 1 January 2005 to 31 December 2015 (Table \@ref(tab:table-icd)).
From early 2016, extensive maintenance and restructuring of the registry has been ongoing, and no new data have been added since 31 December 2015. 

Data from the Primary Care Registry were used in [paper II](#paper2), [paper III](#paper3), [paper IV](#paper4) and [paper VI](#paper6).

### The National Vaccine Registry {#vaccineregistry}

The Icelandic Directorate of Health also maintains the National Vaccine Registry (NVR). 
All vaccine doses administered within the healthcare system are systematically recorded in an individual's electronic health record at the time they are administered.
This record is reviewed and updated regularly, and vaccinations given in other healthcare facilities are included.
The NVR collects this information from all electronic health records in the country.
Vaccines are categorized using the Anatomical-Therapeutic-Chemical (ATC) classification system of the World Health Organization.
All administered vaccine doses with ATC codes "J07AL" (Pneumococcal vaccines) were extracted for the period of 1 January 2005 to 31 December 2017.

Data from the NVR were used in all papers. 

### The National Drug Prescription Registry {#prescriptionregistry}

The national drug prescription registry (NDPR) is a whole-population registry, collected and maintained by the Icelandic Directorate of Health since 1 January 2005
It contains information on all filled drug prescriptions in Iceland.
All pharmacies are required by law to collect data on each filled prescription and submit them to the NDPR.
An important distinction must be made between a written prescription and a filled prescription.
The NDPR receives information if and when a prescription is filled.
It does not record information on written prescriptions that were never filled by the patient.
Therefore, all prescriptions documented within the NDPR were paid for and received by the patient.
Extensive validation and error testing have been performed by the Directorate of Health to ensure the robustness of the NDPR.
Automated electronic submissions, coupled with tightly controlled processes by which pharmacies dispense drugs, has essentially excluded the possibility of any filled prescriptions escaping registration. 

All prescriptions within the ATC therapeutic subgroup "J01" (Antibacterials for Systemic Use), "J07" (Vaccines), "S01" (Opthalmologicals) and "S02" (Otologicals) were extracted for the period from 1 January 2005 to 31 December 2017.
The chemical levels used in the study are shown in Table \@ref(tab:table-atc)

```{r table-atc}
kable(
    x = data.frame(
    stringsAsFactors = FALSE,
    ATC.code = c(
    "J01A",
    "J01B",
    "J01C",
    "J01D",
    "J01E",
    "J01F",
    "J01G",
    "J01M",
    "J01R",
    "J01X",
    "J07A",
    "J07B",
    "J07C",
    "J07X",
    "S01A, S02A",
    "S01C, S02C"
    ),
    Description = c(
    "Tetracyclines",
    "Amphenicols",
    "Beta-lactam antibacterials, penicillins",
    "Other beta-lactam antibacterials",
    "Sulfonamides and trimethoprim",
    "Macrolides, lincosamides and streptogramins",
    "Aminoglycoside antibacterials",
    "Quinolone antibacterials",
    "Combinations of antibacterials",
    "Other antibacterials",
    "Bacterial vaccines",
    "Viral vaccines",
    "Bacterial and viral vaccines",
    "Other vaccines",
    "Anti-infectives",
    "Anti-inflammatory agents and anti-infectives in combination"
    )
    ),
    col.names = c("ATC chemical subgroup code", "Description"),
    caption = "Anatomical Therapeutic Chemical (ATC) codes used for individual-level data collection from the National Drug Prescription Registry. ATC codes are presented down to the therapeutic level, and all sublevels of the listed ATC codes were also included. ATC codes J07 and sublevels were used to extract data on pneumococcal vaccine doses from the National Vaccine Registry."
    )
```

Data from the NDPR were used in [paper II](#paper2), [paper III](#paper3) and [paper IV](#paper4).

### Reimbursement database of Icelandic Health Insurance {#reimbursementdatabase}

The healthcare system in Iceland is a single-payer system with one government-run health insurance provider, under which all permanent citizens are covered. 
Most healthcare visits require a nominal out-of-pocket fee, with the rest of the visit covered by the insurance.
There are exceptions to this -- for example, visits by children under two years of age are completely covered by insurance. 
Healthcare providers are either salaried governmental employees, or independent practitioners who are reimbursed on a per case basis, according to pre-determined negotiations with Icelandic Health Insurance.
To receive pay for services, physicians must submit a reimbursement form, detailing the nature of the visit and any procedures performed using pre-specified procedural codes.
Icelandic Health Insurance maintains a reimbursement database which details the nature and number of procedures performed.
Data on all otolaryngological procedures performed on the middle ear and tonsils were extracted from the reimbursement database for the period from 1 January 2005 to 31 December 2017 using the procedural codes in Table \@ref(tab:table-reimbursement)

```{r table-reimbursement}
kable(
    x = data.frame(
    stringsAsFactors = FALSE,
    reimbursement.code = c(
    "5500601",
    "5500602/55Q0602+55Z0602",
    "5500603/55Q0603+55Z0603",
    "5500604/55Q0604+55Z0604",
    "5501001/55Q1001+55Z1001",
    "5501002/55Q1002+55Z1002",
    "5501201/55Q1201+55Z1201",
    "5501301/55Q1301+55Z1301",
    "5501801/55Q1801+55Z1801",
    "5501802/55Q1802+55Z1802",
    "5501901/55Q1901+55Z1901",
    "5501902/55Q1902+55Z1902",
    "5502002/55Q2002+55Z2002"
    ),
    Description = c(
    "Myringotomy, one or both ears, under local anesthetic",
    "Placement of tympanostomy, one ear (local anesthetic/general anesthesia)",
    "Placement of tympanostomy tube, one ear, and myringotomy, both ears (local anesthetic/general anesthesia)",
    "Removal of tympanostomy tube, one ear (local anesthetic/general anesthesia)",
    "Placement of tympanostomy tube, both ears (local anesthetic/general anesthesia)",
    "Removal of tympanostomy tube, both ears (local anesthetic/general anesthesia)",
    "Adenoidectomy (local anesthetic/general anesthesia)",
    "Adenoidectomy and placement of tymponstomy tube or myringotomy, one or both ears (local anesthetic/general anesthesia)",
    "Tonsillectomy with or without adenoidectomy (local anesthetic/general anesthesia)",
    "Tonsillectomy with or without adenoidectomy - performed with laser (local anesthetic/general anesthesia)",
    "Tonsillectomy, with or without adenoidectomy, and tympanostomy or myringotomy (local anesthetic/general anesthesia)",
    "Tonsillectomy, with or without adenoidectomy, and tympanostomy or myringotomy - performed with laser (local anesthetic/general anesthesia)",
    "Myringoplasty with patch (local anesthetic/general anesthesia)"
    
    )
    ),
    col.names = c("Reimbursement code", "Description"),
    caption = "Reimbursement codes used for individual-level data collection from the Reimbursement database of Icelandic Health Insurance. The codes are specific to Icelandic Health Insurance and do not represent a universal classification system. With one exception (Myringotomy, one or both ears, under local anesthetic), each reimbursable procedure has three associated reimbursement codes. One general (without letters), one specifically for surgeons (Z) and one specifically for anesthesiologists (Q)."
    )
```

Data from the reimbursement database were used in [paper IV](#paper4).
